Supermartingales and Termination Analysis of Statistical PCF

### 2021-11-08 LICS21/antitone.tex
Add more proof in the direction of completeness, and a partial couterexample

\begin{example}[More complex non-affine recursion]
\label{ex:non-affine continuous}
Let
\begin{align*}
\Xi & = \tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample \\
X[e] & = \tif{e \leq p - x^{-2}}{x+1}{f^2(x+1) \oplus_e f(x+1)}
\end{align*}

@@

\begin{therm}
\label{thm:towards completeness}
For any closed term $M$, if there is some function $t : \mathbb N \to [0,1]$ that tends to 1 such that for any $N \in \mathit{Rch}(M)$, the probability that $N$ terminates within $n$ steps is at least $t(n)$, then $M$ is antitone rankable.
\end{therm}

\begin{proof}
\end{proof}

This theorem is not able to account for all antitone rankable terms.

\begin{example}
Let $M = (\tY \lambda f s. \tif{-\tsample < 0}{(\tY \lambda r n. \tif{n < 0}{\underline 0}{r (n - 1)}) s \lfloor \frac 1 \tsample \rfloor}{f (\lambda n. 2^{s n})}) (\lambda n. n)$. It first draws samples repeatedly until one of them is not exactly 0, then choses a random number, applies a function to it, and then loops that many times. Every time the sample drawn is exactly 0, the function gets exponentially faster growing. It therefore doesn't satisfy the conditions of Theorem \ref{thm:towards completeness} because the term reached after $n$ draws of 0 terminates slower as $n$ increases. However, it is antitone rankable nevertheless.
\end{example}

>>>>

\begin{example}[More complex non-affine recursion]
\label{ex:non-affine continuous}
Let
\begin{align*}
\Xi & := \tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample \\
X[e] & := \tif{e \leq p - x^{-2}}{x+1}{f^2(x+1) \oplus_e f(x+1)}
\end{align*}

@@

\begin{therm}
\label{thm:towards completeness}
For any closed term $M$, if there is some function $t : \mathbb N \to [0,1]$ that tends to 1 such that for any $N \in \mathit{Rch}(M)$, there exists some $m$ such that the probability that $N$ terminates within $n$ steps is at least $\frac{t(n+m)-t(m)}{1-t(m)}$, then $M$ is antitone rankable.
\end{therm}

\begin{proof}
The same construction as in the previous lemma applies, with the stopping time $T$ being the $\tY$-runtime $T_M^\tY$, but with $f$ being a concave function $\leq t(n)$ instead of $\leq \mathbb P[T < n]$. The definition of $\epsilon$ is identical, and the definition of the ranking function is similar, but taking the $\tY$-runtime starting at $N$ rather than taking the conditional expectation, so it's $\mathbb E[-\log(1-f(T_N^\tY))]$. While the conditional expectation is only defined up to a set of measure 0, this definition is largely equivalent but works for all $N$. \akr{I feel like it's neater to just leave it at that, and imply the proof that the expectation is bounded is the same as in the more abstract case, but the details are a little different.}

Despite the differences between the constructions, the proofs that the ranking function is finite and satisfies the progress condition are generally similar to the proofs in the case of the abstract stopping time.
\end{proof}

This theorem is not able to account for all terms whose antitone rankability is allowed by the theorems proven so far, as it requires not only that all reachable terms terminate, but that they terminate uniformly quickly. In practice though this is likely to include all reasonable programs.

\begin{example}
The number of functions $\mathbb N \to \mathbb N$ is equal to $|\mathbb R|$, therefore by assigning some code and uncurrying, it is possible to define a function $Z : \mathbb R \times \mathbb N \to \mathbb N$ such that $Z(r, \cdot)$ can be any function $\mathbb N \to \mathbb N$ for some $r \in [0,1]$, and it is even possible to do this in such a way that $Z$ is measurable. As PPCF contains every measurable real function as a primitive, it is therefore possible for a PPCF program to randomly pick a function $\mathbb N \to \mathbb N$ so that every such function is possible.

Let $M = (\lambda r m. (\tY \lambda w n. \tif{n < 0}{\underline 0}{w (n-1)}) \underline Z(r, m)) \tsample \lfloor \frac 1 \tsample \rfloor$. This program first picks a random function, then picks an argument for that function from $\mathbb N$, then counts down from the result to 0. Assuming for convenience that $\frac 1 0$ evaluates to $0$, this clearly terminates almost surely, as do all reachable terms, but for any candidate $t$, there is some $r$ such that $Z(r, \cdot)$ is much faster growing than $t^{-1}$, therefore the condition for Theorem \ref{thm:towards completeness} is not satisfied for $M$. \akr{I don't actually know whether $M$ is antitone rankable. Now I've thought of the example, my guess is that it isn't, which would imply that the completeness conjecture isn't true. If I manage to work it out either way, I'll want to change some of what I've written to account for that, but I'm working for now on the assumption that I won't work it out before publishing.} It is not clear whether this term is antitone rankable at all. However, the construction relies crucially on the peculiar property of PPCF that programs can sample an infinite amount of random data at once (in the form of unlimited precision real numbers). The peculiar function $Z$ also plays a crucial role in the construction, but it would be possible to implement it entirely in terms of basic arithmetic operations, so the fact that it is permitted as a primitive should not be taken as a sign that this is an issue unique to an abstract uncomputable language like PPCF.
\end{example}
